---
title: "Getting started with spatialSVP"
author: "Liubo Li"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with spatialSVP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 5
)
```

```{r setup}
library(spatialSVP)
```

## The idea

Tools such as decoupleR score pathway activity per spot, and tools such as
SpatialDE, SPARK-X, or nnSVG test *genes* for spatial variability. The gap
between the two — *inference on pathway activity maps* — is what spatialSVP
fills. There are two null hypotheses, and
the package is explicit about both:

1. **Absolute null**: "this score vector has no spatial structure at all."
   `test_spatial_variability()` and `permutation_fdr()` answer it with
   shared, block-aware coordinate permutations (three statistic back ends:
   Moran's I, multi-scale covariance, and a Gaussian-process LRT that
   reports a pathway-level *length scale*). On structured tissue this null can
   be false for many ordinary gene aggregates.
2. **Relative (matched) null**: "is this pathway *more* spatially organized
   than random gene sets of the same size and expression level?"
   `matched_null()` + `matched_enrichment()` answer it empirically, without
   coordinate permutations. This is the package's competitive test.

## Simulated data

`simulate_svp_data()` creates spots on a jittered grid where a chosen number
of gene sets carry a spatial signal (Gaussian bump or linear gradient) and
the rest are spatially null:

```{r sim}
sim <- simulate_svp_data(n_spots = 400, n_genes = 160, n_sets = 12,
                         n_spatial_sets = 4, set_size = 10,
                         pattern = "mix", seed = 42)
sim$truth
```

## Score pathway activity

```{r score}
act <- compute_pathway_activity(sim$counts, sim$gene_sets,
                                coords = sim$coords,
                                method = "mean_z",
                                preprocess = "none")
act
```

Any tidy `source/target/mor` network, a named list of gene sets, or a
genes x sets weight matrix works; `decoupler_mlm`, `decoupler_wmean`,
`fgsea`, and `gsva` are alternative scoring back ends. Inputs can also be
`SpatialExperiment` or `Seurat` objects.

## The relative null: matched gene sets

On real tissue, global transcriptomic architecture can make many gene-set
scores spatially structured. An absolute rejection then establishes coordinate
association without showing that an annotated pathway is exceptional.
`matched_null()` draws, for every set, `n_match` random sets of
the same size whose members match the real members one-to-one within
expression bins. Scoring them alongside the real sets and comparing spatial
statistics gives an empirical one-sided p-value per pathway:

```{r matched}
mn <- matched_null(sim$gene_sets, sim$counts, n_match = 25, seed = 1)
mn
act_m <- compute_pathway_activity(sim$counts, mn$network,
                                  coords = sim$coords,
                                  method = "mean_z", preprocess = "none")
enr <- matched_enrichment(act_m, mn, method = "morans_i")
enr[order(enr$p_matched), c("pathway", "stat", "stat_matched_med",
                            "p_matched", "p_adj_BH")]
```

The four planted sets separate from both their own matched replicates
(`stat` vs `stat_matched_med`) and the spatially null sets. With
`n_match = 25` the smallest achievable p-value is 1/26. Resolution rule of
thumb: `n_match >= ~20 * P` is sufficient for a single minimum p-value to
cross a 5% BH threshold (e.g. 50 pathways -> `n_match = 1999`). Fewer
references can suffice when several hypotheses share small p-values; small
`n_match` is mainly useful for screening and ranking.

## Test spatial variability (absolute null)

```{r test-morans}
res_morans <- test_spatial_variability(act, method = "morans_i",
                                       n_perm = 99, seed = 1)
res_morans[order(res_morans$p_value), c("pathway", "stat", "p_value",
                                        "p_adj_BH")]
```

All four planted sets rise to the top. The covariance back end additionally
reports the dominant kernel scale (a coarse descriptor; sharp scale
estimates come from the gp back end):

```{r test-cov}
res_cov <- test_spatial_variability(act, method = "covariance",
                                    n_perm = 99, seed = 1)
res_cov[order(res_cov$p_value), c("pathway", "stat", "p_value",
                                  "dominant_scale")]
```

The Gaussian-process back end reports a per-pathway length scale (in the
units of the coordinates) and the fraction of variance that is spatial.
p-values are permutation-based (`n_perm = 199` by default; use that or more
in practice — the analytic p-value is retained only as an explicit
diagnostic via `n_perm = 0`):

```{r test-gp}
res_gp <- test_spatial_variability(act, method = "gp", n_perm = 99,
                                   seed = 1)
res_gp[order(res_gp$p_value), c("pathway", "p_value", "lengthscale",
                                "spatial_var_fraction")]
```

Short length scales indicate patchy, local programs; long length scales
indicate smooth gradients across the section.

## Multiplicity: shared-permutation Westfall-Young

Pathway scores share genes and therefore correlate strongly.
`permutation_fdr()` shuffles spot labels once per replicate *for all
pathways jointly*, preserving that correlation inside the null, and returns
single-step maxT adjusted p-values:

```{r wy}
fdr <- permutation_fdr(act, method = "morans_i", n_perm = 199, seed = 1)
fdr[order(fdr$p_perm), c("pathway", "stat", "p_perm", "p_adj_BH",
                         "p_adj_wy")]
```

Westfall-Young adjusted p-values dominate the raw permutation p-values by
construction and stay valid under the correlation structure where plain BH
assumptions are shaky. For sharper resolution than the `1/(n_perm+1)`
floor, increase `n_perm` (e.g. 999 on real data).

## Residualization tiers

Under library-size normalization, spatially over-expressed genes claim a
larger share of affected spots' libraries, so unrelated gene sets can
inherit coherent spatial structure (a compositional coupling, not a testing
artifact). `compute_pathway_activity()` offers three tiers:
`residualize = "none"` (default), `"libsize"` (regress each gene's log-CPM
on log10 library size), and `"pcs"` (additionally regress out the top
`n_pcs` expression principal components):

```{r resid}
act_pcs <- compute_pathway_activity(sim$counts, sim$gene_sets,
                                    coords = sim$coords,
                                    method = "mean_z",
                                    residualize = "pcs", n_pcs = 5)
res_pcs <- test_spatial_variability(act_pcs, method = "morans_i",
                                    n_perm = 99, seed = 1)
res_pcs[order(res_pcs$p_value), c("pathway", "stat", "p_value")]
```

The tiers define different estimands. Principal-component regression can remove
a coherent biological pattern when that pattern enters the leading components.
Report total and residual organization separately rather than selecting a tier
from its p-value.

## Visualize

```{r plots, fig.width = 7, fig.height = 5}
plot_pathway_spatial(act, "SPATIAL_1")
```

```{r volcano}
plot_volcano(enr, p_col = "p_matched", fdr_col = "p_adj_BH")
```

```{r lmap}
plot_lengthscale_map(res_gp)
```

## Multi-section / multi-patient designs

Pass `block` (one value per spot, e.g. section or patient id) to
`compute_pathway_activity()`; kernels become block-diagonal and all
permutations shuffle spots only within blocks, avoiding pseudoreplication
when several sections share a coordinate system:

```{r block}
sim_b <- simulate_svp_data(n_spots = 400, n_genes = 160, n_sets = 10,
                           n_spatial_sets = 3, set_size = 10,
                           n_blocks = 2, seed = 7)
act_b <- compute_pathway_activity(sim_b$counts, sim_b$gene_sets,
                                  coords = sim_b$coords, block = sim_b$block,
                                  method = "mean_z", preprocess = "none")
res_b <- test_spatial_variability(act_b, method = "morans_i",
                                  n_perm = 99, seed = 1)
res_b[order(res_b$p_value), c("pathway", "stat", "p_value", "p_adj_BH")]
```

## Session info

```{r session}
sessionInfo()
```
